---
title: "deworm_prophet_time_series"
output: html_document
date: "2023-07-21"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r}
rm(list = ls())
```


```{r}
#import data
df <- read.csv2("cema_internship_task_2023.csv", header=TRUE, sep = ",")
#view the first 3 rows
head(df,3)
```


```{r}
#structure of the dataset
str(df)

```


```{r}
#view number of rows and columns
dim(df)
```
```{r}
#summary of the dataset to get an overview
summary(df) #we notice there are nulls
```


```{r}
#categorize null counts per column
nulls <- colSums(is.na(df))
nulls
```
```{r}
#deal with the nulls
#Assuming the nulls are as a result of no finding you can replace it with 0
df$Acute.Malnutrition <- replace(df$Acute.Malnutrition, is.na(df$Acute.Malnutrition),0)
df$stunted.6.23.months <- replace(df$stunted.6.23.months, is.na(df$stunted.6.23.months),0)
df$stunted.0..6.months <- replace(df$stunted.0..6.months, is.na(df$stunted.0..6.months),0)
df$stunted.24.59.months <- replace(df$stunted.24.59.months , is.na(df$stunted.24.59.months ),0)
summary(df)#no nulls on any column
```


```{r}
#Find total number of children dewormed per county
library(tidyverse)
diahrrea_per_county <- df%>%
  group_by(county)%>%
  summarise(diahrrea_county_count = sum(diarrhoea.cases))%>%
  arrange(desc(diahrrea_county_count))
print(diahrrea_per_county)
```


```{r}
#Deormed per county
dewormed_per_county <- df%>%
  group_by(county)%>%
  summarise(dewormed_county_count = sum(Total.Dewormed))%>%
  arrange(desc(dewormed_county_count))
print(dewormed_per_county)
```


```{r}
malnutrition_per_county <- df%>%
  group_by(county)%>%
  summarise(malnutrition_county_count = sum(Acute.Malnutrition))%>%
  arrange(desc(malnutrition_county_count))
print(malnutrition_per_county)
```


```{r}
#datatype
class(df$period)
```


```{r}
#change period to date format
library(lubridate)
df$period <- parse_date_time(df$period, orders = "%b-%y") #convert period to a proper date format
head(df$period)
class(df$period)
```


```{r}
#to as date
df$period <- as.Date(df$period)
head(df$period)
class(df$period)
```


```{r}
df$yr <- format(df$period, "%Y")
```


##  I HAVE DECIDED TO CHOOSE MY RESEARCH QUESTION TO BE:
        #PREDICTING THE REMAINING ESTIMATED NUMBERS OF DEWORMINGS COUNTRYWIDE
```{r}

#use time series to see total dewormed per year
#Time series
ggplot(data = df, aes(x = period, y= Total.Dewormed, group = county, color = county ))+
  geom_line()+
  theme_minimal()
```


```{r}
#separate years in total deworms
deworm_per_yr <- df%>%
  group_by(county, yr)%>%
  summarise(Total_deworm_per_yr = sum(Total.Dewormed))
print(deworm_per_yr)
```


```{r}
#highest dewormings by county
highest_deworms <- deworm_per_yr%>%
  group_by(county)%>%
  filter(Total_deworm_per_yr == max(Total_deworm_per_yr))
print(highest_deworms)
```


```{r}
#year with most occurences in Kenya
yr_count <- table(highest_deworms$yr)
most_frequent <- names(yr_count)[which.max(yr_count)]
yr_count
print(paste("Year with most dewormings done in Kenya:",most_frequent))
```


```{r}
#year with most deworming done with total number in Kenya
total_most_dewormig <- df%>%
  group_by(yr)%>%
  summarise(Total_deworm_per_yr = sum(Total.Dewormed))
most_deworming_year <- total_most_dewormig%>%
  filter(Total_deworm_per_yr == max(Total_deworm_per_yr))
most_deworming_year
```


```{r}
#let us try to predict deworming activities for the rest of the year viafacebook prophet
#filter data up to the last month 

library(prophet)
library(tidyverse)


end_date <-as.Date("2023-06-01")
start_date <- end_date - months(6)
```


```{r}
#since we are predicting for each month for the remaining part of the year
df$year_month <- format(df$period, "%Y-%m-%d")

past_6_months <- df%>%
  filter(period >=start_date & period <= end_date)


total_past_6_months <- past_6_months%>%
  group_by(year_month)%>%
  summarise(Total_dewormed = sum(Total.Dewormed))
print(total_past_6_months)
```
```{r}
#rename columns according to prophet time series
prophet_data <- total_past_6_months%>%
  rename(ds=year_month, y=Total_dewormed)
print(prophet_data)

prophet_data$ds <- as.Date(prophet_data$ds, format = "%Y-%m-%d")
class(prophet_data$ds)
```


```{r}

#fit the model
prophet_model <- prophet(prophet_data)
```


```{r}
#dataframe with July to Dec 2023 dates
future_dates <- data.frame(ds = seq(as.Date("2023-07-01"), as.Date("2023-12-01"), by = "month"))
predicted_numbers <- predict(prophet_model, future_dates)
plot(prophet_model, predicted_numbers, main = "Predictions for the remainder of the year")
```


```{r}
#show them per month. Countrywide
predicted_numbers_monthly <- predicted_numbers%>%
  group_by(ds)%>%
  summarise(predicter_values = sum(yhat))
print(predicted_numbers_monthly)
```



##This will help in the preparation for the acquisition of both required resources and manpower
We can go ahead to find distributions across the counties for the remaining part of the year since we have both dataframes with deworming figures on each county and predicted figures.
Thank you









